---
title: "Statistical Inference: Simulation Exercise"
author: "Mitch Murphy"
date: "4/17/2018"
output:
  pdf_document: default
---
## Introduction 
1. Simulate the exponential distribution
  1. Calculate the mean and variability
  2. Exploratory Data Analysis
    + Highlight basic features present in the data
  3. Construct and interpret confidence intervals
  4. What assumptions must hold?
 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Simulation
Let us first sum up the exponential distribution: $\mu=\frac{1}{\lambda}$ and a variance  $\sigma^2=\lambda^{-2}$. 

```{r}
set.seed(round(1000*pi, 0))

lambda <- 0.2
# We willl be taking the average of 40 exponentials
n <- 40
# we will be comparing this distribution to 1000 random uniforms
numOfSims <- 1000
simulatedExponentials <- replicate(numOfSims, rexp(n, lambda))
# now lets compute the mean of these simulations
means <- apply(simulatedExponentials, 2, mean)
d <- density(means)
```

## Mean
Recall that $\mu=\frac{1}{\lambda}$ so theoretically $\mu=5$ and our sample comes close, $\mu=0.5025$ in this case. 
```{r mean}
plot(d, main = "Kernel Density of Means", xlab = "Mean")
polygon(d, col = "red", border = "blue")
observed_mean <- mean(means)
legend("topleft", legend = c(paste("Observed Mean:",round(observed_mean,3)), paste("Expected Mean:",lambda^-1)),
       lty = c(2,3), col = c("yellow", "black"), bty = "n")
abline(v=lambda^-1, col = "yellow", lwd = 2, lty = 2)
abline(v=observed_mean, col = "black", lty = 3, lwd = 2)
```

## Variability 
In our case we must adjust $\sigma$ for the given sample size ($n=40$) so we have 
$\sigma^2=(\frac{1}{\lambda}\cdot\frac{1}{\sqrt{n}})^2$. Theoretically $\sigma^2=0.625$, however in this case
$\sigma^2=0.6624831$

```{r variability}
# first let's compute the standard deviations
sample_sd <- sd(means)
expected_sd <- (1/lambda)/sqrt(n)
# now let's focus on the variance of the 2
sample_var <- sample_sd^2
expected_var <- ((1/lambda)*(1/sqrt(n)))^2
compareMatrix <- matrix(c(1/lambda,observed_mean,
                          expected_var, sample_var), nrow = 2)
rownames(compareMatrix) <- c("Expected","Observed")
colnames(compareMatrix) <- c("mean", "variance")
compareMatrix
```

## Normality
```{r normality}
# lets overlay the normal distribution over our observed data (density/histogram)
xfit <- seq(min(means), max(means), length=100)
yfit <- dnorm(xfit, mean=1/lambda, sd=(1/lambda/sqrt(n)))
hist(means,breaks=n,prob=T,col="orange",xlab = "means",main="Density of means",ylab="density")
lines(xfit, yfit, pch=22, col="black", lty=5)

qqnorm(means)
qqline(means, col = 2)
```
The distribution of the averages of 40 exponentials appears to be nearly identical to that of the normal distribution. 

# Assumptions
When constructing confidence intervals (as we did above), we assumed that the variance of the 2 groups was different.