---
title: 'Statistical Inference: Tooth Growth'
author: "Mitch Murphy"
date: "4/17/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(datasets)
library(ggplot2)
```

# Tooth Growth
## Summary of Data
Now let us take a look at the affect of two different supplements on tooth growth. First, let's get an high-level overview
of the data:
```{r toothGrowth}
tg <- ToothGrowth
tg$dose <- as.factor(tg$dose)
str(tg)
summary(tg)

ggplot(data=tg, aes(x=dose, y=len, fill=supp)) + geom_bar(stat="identity",) + facet_grid(. ~ supp) + 
  xlab("Dose, mg") + ylab("Tooth length") + guides(fill=guide_legend(title="Supplement"))
```

## Confidence Intervals
Now let us construct 95% confidence intervals across the different doses.

```{r}
ci <- t.test(len ~ supp, data = tg)
pval <- ci$p.value
ci
```
Since zero is contained in this interval, and because p-val$\geq \alpha=0.05$ we cannot reject the null hypothesis.
Let's further explore this data by creating dosage-groups, as so:
```{r}
tg.dose1 <- subset(tg, dose %in% c(0.5,1.0))
tg_dose2 <- subset(tg, dose %in% c(0.5,2.0))
tg_dose3 <- subset(tg, dose %in% c(1.0,2.0))

# construct the CI for dose of 0.5 and 1.0
t.test(len ~ dose, data = tg.dose1)
#construct the CI for a dose of 0.5 and 2.0
t.test(len ~ dose, data = tg_dose2)
# construct the CI for a dose of 1.0 and 2.0
t.test(len ~ dose, data = tg_dose3)
```
Because zero is not contained in any of the above confidence intervals and the p-value $\leq \alpha$ we can conclude
that the dose of the supplement effects tooth length; specifically as the dose increases so does the tooth length.
We have already concluded that the supplement type is not significant (first CI above).

# Assumptions
When constructing confidence intervals (as we did above), we assumed that the variance of the 2 groups was different.